---
title: "Spurious correlations"
author: "Arnaud Legrand"
date: "12/8/2021"
output: html_document
---

# Known regressors

From https://statisticsbyjim.com/regression/multicollinearity-in-regression-analysis/
- Structural multicollinearity: This type occurs when we create a model term using other terms. In other words, it’s a byproduct of the model that we specify rather than being present in the data itself. For example, if you square term X to model curvature, clearly there is a correlation between X and X2.
- Data multicollinearity: This type of multicollinearity is present in the data itself rather than being an artifact of our model. Observational experiments are more likely to exhibit this kind of multicollinearity.

Here is an illustration.
```{r}
    library(tidyverse)    
    
    sim <- function(rho){
      #Number of samples to draw
      N = 50
    
      #Make a covariance matrix
      covar = matrix(c(1,rho, rho, 1), byrow = T, nrow = 2)

      # Append a column of 1s to N draws from a 2-dimensional  
      # Gaussian 
      # With covariance matrix covar
      X = cbind(rep(1,N),MASS::mvrnorm(N, mu = c(0,0), 
                  Sigma = covar))

      # True betas for our regression
      betas = c(1,2,4)

      # Make the outcome
      y = X%*%betas + rnorm(N,0,1)

      # Fit a linear model
      model = lm(y ~ X[,2] + X[,3])
      
      # Return a dataframe of the coefficients
      return(tibble(a1 = coef(model)[2], a2 = coef(model)[3]))     
    }
    
    #Run the function 1000 times and stack the results
    zero_covar = rerun(1000, sim(0)) %>% 
                 bind_rows
    
    #Same as above, but the covariance in covar matrix 
    #is now non-zero
    high_covar = rerun(1000, sim(0.95)) %>% bind_rows
    
    #plot
    zero_covar %>% 
      ggplot(aes(a1,a2)) +
      geom_point(data = high_covar, color = 'red') +
      geom_point()

```

# Error in variables
## One variable
```{r}
a = 5
b = 2
N = 1000
X = runif(n = N, min = -2, max = 2)
Xa = X + rnorm(n = N, sd=.2)
Y = a*X + b + rnorm(n = N)
df = data.frame(Y=Y, X=X, Xa=Xa)

summary(lm(data=df, Y~X))
summary(lm(data=df, Y~Xa))
```

## Two variables
From "Inflation of Type I error in multiple regression when independent variables are measured with error"
http://www.utstat.toronto.edu/~brunner/MeasurementError/MeasurementError7b.pdf 
This article quantifies this kind of effect:
https://journals.plos.org/plosone/article?id=10.1371/journal.pone.0152719

Imagine that you try to explain the number of daily swimming pool deaths (Y) from the number of ice cream cones sold per days (X1). Obviously, temperature plays a role so you may be tempted to add the subjective heat (X2) in the picture to check whether ice cones have a real effect or not. Unfortunately, both are observations, i.e., what you care about is the number of ice cream cones consumed before going to the pool per day (Xi1) and the real temperature (Xi2), and Xi1 and Xi2 are obviously correlated variables.

```{r}
library(MASS)
library(dplyr)
b0 = 2
b1 = 3
b2 = 0
N = 500

mu = c(3,10)
Sigma = matrix(data=c(1,0.8,0.8,2), nrow = 2)
Xi = mvrnorm(n = N, mu, Sigma)

nu = c(0,0)
Theta = matrix(data=c(.3,-0.1,-0.1,.3), nrow = 2)
X = Xi + mvrnorm(n = N, nu, Theta)

df = as.data.frame(cbind(Xi,X))
names(df)=c("Xi1","Xi2","X1","X2")
plot(df)

df %>% mutate(Y = b0 + b1*Xi1 + b2*Xi2 + rnorm(n=n())) -> df

plot(df[c("Y","Xi1","Xi2")])
plot(df[c("Y","X1","X2")])

summary(lm(data=df, Y~Xi1+Xi2))
# summary(lm(data=df, Y~X1+X2))
```

# Spurious correlation
```{r}
library(ggplot2)
spurious_correlation = function(N = 10) {
  trial = 1
  while(TRUE) {
      X = runif(n = N, min = 5, max = 6)
      Y = runif(n = N, min = 20, max = 70)
      if(abs(cor(x = X, y = Y))>.95) {
          print(trial)
          return(data.frame(x = X, y = Y))
      }
      trial = trial + 1
  }
}
d = spurious_correlation();
d
summary(d)
cor(d)
ggplot(data = d, aes(x=x,y=y)) + geom_point() + geom_smooth(method="lm")

```

