---
title: "French given names per year per department"
author: "Lucas Mello Schnorr, Jean-Marc Vincent"
date: "October, 2022"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# The problem context
The aim of the activity is to develop a methodology to answer a specific question on a given dataset. 

The dataset is the set of Firstname given in France on a large period of time. 
[https://www.insee.fr/fr/statistiques/2540004](https://www.insee.fr/fr/statistiques/fichier/2540004/dpt2021_csv.zip), we choose this dataset because it is sufficiently large, you can't do the analysis by hand, the structure is simple


You need to use the _tidyverse_ for this analysis. Unzip the file _dpt2020_txt.zip_ (to get the **dpt2020.csv**). Read in R with this code. Note that you might need to install the `readr` package with the appropriate command.

## Download Raw Data from the website
```{r}
file = "dpt2021_csv.zip"
if(!file.exists(file)){
  download.file("https://www.insee.fr/fr/statistiques/fichier/2540004/dpt2021_csv.zip",
	destfile=file)
}
unzip(file)
```
Check if your file is the same as in the first analysis (reproducibility)
```{bash}
md5 dpt2021.csv
```
expected :
MD5 (dpt2021.csv) = f18a7d627883a0b248a0d59374f3bab7

## Build the Dataframe from file

```{r}
library(tidyverse)

FirstNames <- read_delim("dpt2021.csv",delim=";")
```

All of these following questions may need a preliminary analysis of the data, feel free to present answers and justifications in your own order and structure your report as it should be for a scientific report.

1. Choose a firstname and analyse its frequency along time. Compare several firstnames frequency
2. Establish by gender the most given firstname by year. Analyse the evolution of the most frequent firstname.
3. Optional : Which department has a larger variety of names along time ? Is there some sort of geographical correlation with the data?
