---
title: "R crash course for Computer Scientists"
author: "Arnaud Legrand"
date: '2022-10-06'
output: 
  html_document:
    toc: true
    toc_depth: 4
    toc_float: true
    theme: united
    highlight: tango
---


## R is a high level functional and object oriented language

Keep in mind you should not consider R as a classical generic programming language. It is particularly well suited to data analysis involving statistics and probabilities. 

### Object orientation
R provides a few built-in constants, variables, and functions to play with.

```{r}
pi
letters
iris
mean
```

Every variable is actually object so you can get information through `str` (structure) and `typeof`. 
```{r}
str(pi); typeof(pi)
str(letters); typeof(letters)
str(mtcars); typeof(mtcars)
str(mean); typeof(mean)
```
You can access the documentation with `?`
```{r}
?mean
```

### High level data structures
Basic types are `num`, `int`, `char`, and factors (e.g., Male or Female, like an `enum` in C) but R also provides several built-in higher-level datatypes:
- Arrays or vectors with all elements being of the same type
- Lists: collections of named objects, some kind of hash table
- Matrices: 2D arrays
- Data frames: 

The types I use the more are vectors and data frames.

Let's build a few ones.
```{r}
V = c(1,5,12) # c : Combine Values into a Vector or List
V
str(V)
V[1]  # Index starts at 1!
V[4] = 15 ; V
length(V)
V[0]  # Uh ?
V[-1] # Oh... ok, this removes the first element
V[5]  # NA means "Not Available"
V = c(V,-5,1:4, seq.int(from=2, to = 10, by = 3), rep_len(NA,4), V);V
```
So you'd better be careful with indices. Array are dynamically resized but it comes with a cost so if you know the size in advance, you'd better initialize them.

Data frames are very handy and can be seen as a list of vectors which have all the same size.
```{r}
str(iris)
dim(iris)
iris[3,1]
iris[3,"Sepal.Length"]
iris$Sepal.Length
iris$Sepal.Length[3] # much more convenient
df = data.frame(Col1 = V, Col2 = "Hello", Col3 = 1:2, Col4 = as.factor(c("A", "B", "C","D")))
str(df)
df
```
You can also combine data frames together with `rbind` and `cbind`:
```{r}
nrow(df)
df = rbind(df,df)
nrow(df)
```
But in practice, you often read dataframes from existing databases or datafiles, e.g., with `read.csv`.
```{r}
df = read.csv("https://www.stats.govt.nz/assets/Uploads/Consumers-price-index/Consumers-price-index-September-2022-quarter/Download-data/consumers-price-index-september-2022-quarter-seasonally-adjusted.csv", header = T, sep=",")
str(df)
head(df)
```


### Vectorization
Obviously, R provides classical imperative control structures like branching, `for` and `while` loops but there is generally a better style.
So avoid them as much as possible.

```{r}
for(i in 1:length(V)) {
  print(paste0("V[",i,"]^2 = ",V[i]^2))
}
```
Note that you may wonder yourself how tests react to `NA` ?...

Anyway, R is a high level language which provides *vectorization*, i.e. most built-in operators and functions handle seamlessly vectors.
```{r}
V = 1:5
V+12
V = log(V^3) + 3*sin(V); V
V <= 3
```
Even subsetting is vectorized
```{r}
V[c(5,1:2)]
V[V <= 2]
```
This is particularly convenient with dataframes as you can easily extract information "à la SQL".
```{r}
d = iris[iris$Sepal.Length <=mean(iris$Sepal.Length) & 
         iris$Petal.Length <=mean(iris$Petal.Length) ,]
d
d$Weird = log(d$Sepal.Length)*sin(d$Sepal.Width) 
summary(d)
```
So R being a functional language, it provides many simple functions, which, once well combined are very powerful. Check for example `sum`, `cumsum`, `ifthen`, ...

### Random number generation

R embeds every random generator you can dream of. The naming convention is as follows for a law `law`:
- `rlaw`: random number generator for `law`
- `dlaw`: density function ($\mathbb{R}\to\mathbb{R}^+$)
- `plaw`: cumulative density function ($\mathbb{R}\to[0,1]$)
- `qlaw`: reciprocal of the cdf, i.e., quantile ($[0,1]\to\mathbb{R}$)
Classical laws are `unif`, `norm`, `exp`, `cauchy`, `beta`, ...
```{r}
rnorm(n = 10, mean = 2, sd = 1)
```
For custom discrete distributions, you may want to use the `sample` function.

### Statistics and plots
Obviously, you have access to `mean`, `min`, `max`, `median`, `var`, `sd`, `summary` (which properly handle `NA`s) but also tests `chisq.test`, `wilcox.test`, etc. The default choices are generally good choices from a statistician point of view and formulas are correct (e.g., unbiased estimators).

Default plots are not very nice but quite effective.
```{r}
plot(rnorm(1000), rnorm(1000))
hist(rnorm(1000))
plot.ts(runif(100))
x = rexp(100)
qqnorm(x)
plot(iris)
```



### Function
As explained earlier, there are many built-in functions but you may want to write your own. In R, functions have named arguments with possible default value (so it acts as optional arguments).

```{r}
mynorm2 = function(n = 10, Mean = 2, SD = 1.2, power = 2) { # Uggly variable names. Do not do this!
  return(rnorm(n = n, mean = Mean, sd = SD)^power)
}
mynorm2() # What does this do ? 
mynorm2(5,-1,1.1,10) # Err ?
mynorm2(Mean = 2 , power = 1.4, SD = 1.4) # Better
```

### R can be quite surprising
Lexical scoping. There are several assignment operators. AFAIK `=` and `<-` (or `->`) are equivalent but `<<-` could be what you're looking for if you need side effects in recursive functions... :)
```{r}
x <- 1
f = function() {
  x <- 5 # or 5 -> x
  print(x)
}
f()
print(x)
g = function() {
  x <<- 5
  print(x)
}
g()
print(x)
```
Here is an other interesting feature from a semantic aspect. How the hell does R know the (outer) variable names inside the function ? :)
```{r}
plot(mtcars$mpg,mtcars$qsec)
```

## The tidyverse

### Dplyr
The R function names and API is not very consistent nor convenient when working with data frames. This is quite error prone.
```{r}
mtcars[mtcars$cyl==6 & mtcars$hp >120,]
```

```{r}
library(dplyr)
```


```{r}
mtcars %>% filter(cyl==6 & hp >120)
```
```{r}
mtcars %>% filter(cyl==6 & hp >120) %>% select(mpg, cyl, hp, qsec)
```
```{r}
mtcars %>% group_by(cyl,vs) %>% summarise(qsec_mean=mean(qsec),
                                       qsec_sd=sd(qsec),
                                       num = n()) -> df
```
Efficient, expressive, avoids side effects. Can be used to create splitted models. See R4DS for more.

### Ggplot

R plots can be surprising, uggly, and have funky documentations with obscure arguments.
```{r}
plot(mtcars)
```

```{r}
plot(mtcars$qsec,mtcars$disp)
```

But ggplot2 empowers you with lego blocks and a high level grammar of graphics
```{r}
library(ggplot2)
ggplot(mtcars, aes(x = mpg, y = drat )) + geom_point() + 
  geom_point(data = cars, aes(x = speed, y=dist), color = "red") + 
  theme_bw() + geom_smooth(method = "lm")
```

